A Short and Mechanized Consistency Proof for Dependent Type Theory

### 2023-12-25 paper/paper.tex
Add the inversion lemmas

\section{Semantic Typing}
\label{sec:logreldep}
\begin{figure}[h]
  \begin{equation*}
    \begin{split}
      [-]_= &:: \mathit{Term} \rightarrow \mathit{Nat} \rightarrow \mathcal{P}(\mathit{Term})\\
      [ [[Void]] ]_i &= \emptyset \\
      [ [[Bool]] ]_i &= \{ [[a]] \mid [[a =>+ true]] \text{ or } [[a =>+ false]] \} \\
      [ [[a ~ b : A]] ]_i &= \{ [[p]] \mid [[p =>+ refl]] \text{ and } [[a<=>b]] \}\\
      [ [[Pi A B]] ]_i &= \{ [[b]] \mid \forall [[a]], \text{ if } [[a]] \in [ [[A]] ]_i\text{, then } [[b a]] \in [ [[B {a}]] ]_i \} \\
      [ [[Set j]] ]_i &= \{ [[A]] \mid [[j < i]] \text{ and } [[A]] \in dom([-]_j) \}\\
      [ [[ A ]] ]_i &= \{ [[a]] \mid \exists [[B]], [[a]] \in[ [[B]] ]_i \text{ and } [[A => B]] \}
    \end{split}
  \end{equation*}
  \caption{Specification of the logical relation for \lang}
  \label{fig:logrel}
\end{figure}

In this section, we present the definition of the logical relation as a
partial function.
Figure~\ref{fig:logrel} shows the specification
of the logical relation that we are about to define. The specification
itself does not immediately induce a definition as it is unclear
whether there exists a smallest partial function that satisfies the
specification.

The construction of a function that meets the specification in
Figure~\ref{fig:logrel} can be broken down into two steps. To build an
infinite hierarchy of universes, we first define a recursor that constructs
the interpretation of the universe level $[[n]]$ given a parameterized
interpretation function for universe levels that are strictly less
than $[[n]]$.
Next, we use the well-foundedness of the less than relation to tie the
knot and define the full interpretation function that meets the
specification.
The resulting definition is then shown to be functional and thus
justifies the notation $[ [[A]] ]_i$ since we know there would be an
unambiguous output corresponding to the type $[[A]]$, assuming that
$[[A]]$ is in the domain of the partial function.

>>>>

\section{Semantic Typing}
\label{sec:logreldep}
\begin{figure}[h]
\drules[I]{$[[Interp I i A S]]$}{Logical Relation}{Void, Bool, Eq, Pi, Set, Red}
\caption{Logical relation for \lang}
\label{fig:logrel}
\end{figure}
In this section, we define the logical relation for \lang{} in the
form of an inductively defined relation (Figure~\ref{fig:logrel}). The
logical relation takes the form $[[Interp I i A S]]$. The
metavariables $[[A]]$ and $[[i]]$ stand for terms and natural
numbers respectively, as introduced earlier in
Figure~\ref{fig:syntax}.
The metavariables $[[I]]$ and $[[S]]$ are
sets with the following signatures:
\begin{equation*}
  \begin{split}
    [[I]] &\in [[ { j | j < i  } ->  PowerSet STm ]] \\
    [[S]] &\in [[PowerSet STm]]
  \end{split}
\end{equation*}
The function $[[I]]$ is a family of subsets of terms indexed by
natural numbers strictly less than the parameter $[[i]]$, which
represents the current universe level.  In \rref{I-Set}, function
$[[I]]$ is used to define the meaning of
universes that are strictly smaller than the current level $[[i]]$. The
restriction $[[j < i]]$ in \rref{I-Set} is crucial for our system to
be predicative. Removing the ordering constraint would result in a
system where one can encode Girard's paradox~\citep{girard-thesis}.

\begin{figure}[h]
\begin{equation*}
  \begin{split}
    [[InterpR i A S]] &:= [[ Interp F i A S  ]] \\
    [[F i]] &:= [[exists S , InterpR i A S]]
  \end{split}
\end{equation*}
\caption{Logical relation for all universe levels}
\label{fig:logrelrec}
\end{figure}
To tie the knot and obtain an interpretation of all universe levels,
we define in Figure~\ref{fig:logrelrec} the final version of our interpretation judgment recursively
using the well-foundedness of the strict less than relation on natural
numbers.
The judgment $[[InterpR i A S]]$ now reads that the type $[[A]]$ is a
level-$[[i]]$ type \emph{semantically} inhabited by terms from the set
$[[S]]$. The way it's constructed enables us to reason about its
property by nested induction over the less than relation and the
structure of the relation $[[Interp I i A S]]$, though for the
majority of the cases we do not need any information about the
parameterized function $[[I]]$, where the property about $[[InterpR i
A S]]$ follows as a corollary of a property about $[[Interp I i A S]]$
with no assumptions imposed on $[[I]]$.

For the rest of the section, we establish some important facts about
the logical relation that will be useful for proving fundamental lemma
in Section~\ref{sec:logrelproof}.

The relation $[[Interp I i A S]]$ satisfies the following inversion
principles.
\begin{lemma}[Void Inv]
  \label{lemma:voidinv}
  If $[[Interp I i Void S]]$, then $[[S = emptyset]]$.
\end{lemma}
\begin{lemma}[Bool Inv]
  \label{lemma:boolinv}
  If $[[Interp I i Bool S]]$, then $[[S = { a | a =>+ true \/ a =>+ false   }]]$.
\end{lemma}
\begin{lemma}[Eq Inv]
  \label{lemma:eqinv}
  If $[[Interp I i a ~ b : A S]]$, then $[[S = { p | p =>+ refl , a <=> b  }]]$.
\end{lemma}
\begin{lemma}[Pi Inv]
  \label{lemma:piinv}
  If $[[Interp I i Pi A B S]]$, then there exists $[[S]],[[F]]$ such
  that the following conditions hold:
  \begin{itemize}
  \item $[[Interp I i A S ]]$
  \item $[[F in S -> PowerSet STm]]$
  \item $[[forall a, (# a in S implies Interp I i B { a } F a #)]]$
  \end{itemize}
\end{lemma}
\begin{lemma}
  \label{lemma:setinv}
  If $[[Interp I i Set j S]]$, then $[[j < i]]$ and $[[S = I j]]$.
\end{lemma}
